---
title: "2 - Probability theory and statistics review"
output: html_notebook
---

# Basic probability theory

```{r setup}
library(tidyverse)
```


## Introductory definitions

* *Random process*: A process that, when repeated multiple times, can lead to different outcomes. Can be _discrete_ or _continuous_.
* *Sample space*: Set of all possible outcomes a random process can have.
* *Logical independence*: Two events that can not affect one another.
* *Statistical independence*: Two events $A$ and $B$ are statistically independent if $Pr(A|B) = Pr(A)$.
* *Joint probability*: For independent events, calculating joint probabilities is to multiply the maringal probabilities: $Pr(A, B) = Pr(A)Pr(B)$. For example: $Pr(heads, heads, heads) = (.5)^3$.

> Question: What is the probability of rolling a 7 using two dice, and is it the same as the probability of rolling a 3?

```{r}
df <- crossing(D1 = 1:6, D2 = 1:6) %>% 
  mutate(outcome = D1 + D2)

tibble(
  "Pr(outcome = 7)" = df %>% 
    filter(outcome == 7) %>% 
    nrow(),
  "Pr(outcome = 3)" = df %>% 
    filter(outcome == 3) %>% 
    nrow()) %>% 
  magrittr::divide_by(nrow(df))
```

# Probability tree

# Contingency tables

# Monty hall example

# Exercises
